Calendar-based graphics for visualising people's daily schedules

### 2017-07-11 main.Rmd
A few tweaks

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualising people's daily schedules"
  plain: "Calendar-based graphics for visualising people's daily schedules"
abstract: >
  It is challenging to visualise people's daily schedules as such data are often of multiple seasons and of long time span. One example is Melbourne pedestrian sensor data which are collected at hourly interval and at a number of locations. Owing to human activities, the data features multiple temporal scales such as time of day, day of week, day of year. We develop the \code{frame\_calendar} function that provides a method for rearranging such kind of data in a calendar-based format and accomplishing visualisation with the grammar of graphics, and describe its construction in detail. We provide some examples to demonstrate its usage by applying to the pedetrian data.
keywords:
  # at least one keyword must be supplied
  formatted: [calendar-based visualisation, temporal-context data]
  plain:     [calendar-based visualisation, temporal-context data]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
  citation_package: biblatex
bibliography: ["references.bib", "packages.bib"]
---

@@

This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 installed sensors counting pedestrians every hour across the downtown area until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed lights into understanding people's daily schedules, or assisting administration and business planning. We start off with the conventional time series plot to take a glimpse at such data. Small multiples, shown as Figure \ref{fig:time-series-plot}, give an overall picture of foot traffic at a selection of 7 sensors. Figure \ref{fig:facet-time}, complemented to the Figure \ref{fig:time-series-plot}, provides another aspect of temporal patterns in the more detailed level by faceting day of the week. It lends itself to a number of issues that make exploratory data visualisation challenging in many temporal-context applications involving human behaviours:

@@

```{r time-series-plot, fig.height = 6, fig.cap = "Time series plot about the number of pedestrians over the year of 2016 at a selection of 7 sensors in the city of Melbourne. Coloured by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. The traffic, specifically at two train stations (Flagstaff Station and Flinders Street Station), is primarily driven by the commuters, as depicted in the graph."}
```

@@

Calendar-based graphics turn out to be a useful tool in unfolding human-related activities in temporal context (cite). For example, @VanWijkCluster1999 developed a calendar view of heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It remarkably contrasts weekdays and weekends, highlights public holiday, and presents other known seasons like school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in a couple of R packages: \pkg{ggTimeSeries} [@R-ggTimeSeries] and \pkg{ggcal} [@R-ggcal]. However, these techniques are too constrained to colour-encoding graphics and day of week as the smallest time scale. Time of day, which serves as one of the most important aspect in explaining variations arising from pedestrian sensor data, will be neglected through daily aggregation. Additionally if simply using colours encoded as quantities, it may becomes less perceivable of the actual variations or behaviours from the viewer's perspective.

@@

The remainder of the paper is organised as follows. Section \ref{sec:algorithm} describes the \code{frame_calendar} function and its construction. Section \ref{sec:examples} presents some examples of its usage. Section \ref{sec:discussion} discusses the advantages and disadvantages of the method.

@@

Figure \ref{fig:flinders-2016} shows the line glyphs organised in the monthly calendar over the year of 2016. This is achieved by the \code{frame_calendar} function computing the new coordinates according to the input data variables; in turn the rearranged data values are plotted using the \pkg{ggplot2} package [@R-ggplot2], which is the implementation of the grammar of graphics (@wilkinson2006grammar and @wickham2010layered). The \code{frame_calendar} function assumes that the underlying data structure is tidy data [@wickham2014tidy].

<!-- ``` -->
<!-- frame_calendar( -->
<!--   data, x, y, date, calendar = "monthly", dir = "h", sunday = FALSE,  -->
<!--   nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed" -->
<!-- ) -->
<!-- ``` -->

@@

# Options

@@

The comparison of one sensor to the other adds additional insights to the dataset, which is commonly done with an overlaying plot such as Figure \ref{fig:overlay}. For instance, the dominant patterns occurred to both train stations--Flinders Street Station and Flagstaff station--are together driven by the commuters on the work days; however, the former is overwhelmed by the number of pedestrians during the weekends and public holidays, compared to the latter. It suggests that Flagstaff station limits its functionality as simply the public transport hub for day-to-day commuters, but various activities take place around the Flinders Street Station other than commuting. Because Flinders Street Station closely lies to South Bank and other places of attractions where the locals go for leisure and the tourists for travelling.

```{r overlay, fig.height = 6.5, fig.cap = "Superimposing line graphs of two sensors on the monthly calendar. The orange line, represented by the number of pedestrians at Flinders Street Station, is on top of the green one indicating Flagstaff Station. Flinders Street Station attracts more pedestrians during the afternoon time and non-working days."}
```

The nighttime events, such as White Night and New Year Evening Firework, has barely affected the operation of Flagstaff Station but heavily affected the incoming traffic to Flinders Street Station on these days.

@@

```{r scatterplot, fig.height = 5, fig.cap = "Lag scatterplot on the daily calendar. Lagged hourly counts are plotted against current ones to demonstrate the autocorrelation at the order of 1. The positive association between the one-hour-ahead count and the present regardless of the data values takes place on non-working days."}
```

>>>>

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualising people's daily schedules"
  plain: "Calendar-based graphics for visualising people's daily schedules"
abstract: >
  It is challenging to visualise people's daily schedules as such data are often of multiple seasons and of long time span. One example is Melbourne pedestrian sensor data which are collected at hourly interval and at a number of locations. Owing to human activities, the data features multiple temporal scales such as time of day, day of week, day of year. We develop the \code{frame\_calendar} function that provides a method for rearranging such kind of data in a calendar-based format and accomplishing visualisation with the grammar of graphics. We describe its construction along with a few options. We provide some examples to demonstrate its usage by applying to the pedetrian data.
keywords:
  # at least one keyword must be supplied
  formatted: [calendar-based visualisation, temporal-context data]
  plain:     [calendar-based visualisation, temporal-context data]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
  citation_package: biblatex
bibliography: ["references.bib", "packages.bib"]
---

@@

This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 installed sensors counting pedestrians every hour across the downtown area until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed lights into understanding people's daily schedules, or assisting administration and business planning. We start off with the conventional time series plot to take a glimpse at such data. Small multiples, shown as Figure \ref{fig:time-series-plot}, give an overall picture of foot traffic at a selection of 3 sensors. Figure \ref{fig:facet-time}, complemented to the Figure \ref{fig:time-series-plot}, provides another aspect of temporal patterns in the more detailed level by faceting day of the week. It lends itself to a number of issues that make exploratory data visualisation challenging in many temporal-context applications involving human behaviours:

@@

```{r time-series-plot, fig.height = 6, fig.cap = "Time series plot about the number of pedestrians over the year of 2016 at a selection of 3 sensors in the city of Melbourne. Coloured by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. The traffic, specifically at two train stations (Flagstaff Station and Flinders Street Station), is primarily driven by the commuters, as depicted in the graph."}
```

@@

Calendar-based graphics turn out to be a useful tool in unfolding human-related activities in temporal context. For example, @VanWijkCluster1999 developed a calendar view of heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It remarkably contrasts weekdays and weekends, highlights public holiday, and presents other known seasons like school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in a couple of R packages: \pkg{ggTimeSeries} [@R-ggTimeSeries] and \pkg{ggcal} [@R-ggcal]. However, these techniques are too constrained to colour-encoding graphics and day of week as the smallest time scale. Time of day, which serves as one of the most important aspect in explaining variations arising from pedestrian sensor data, will be neglected through daily aggregation. Additionally if simply using colours encoded as quantities, it may becomes less perceivable of the actual variations or behaviours from the viewer's perspective.

@@

The remainder of the paper is organised as follows. Section \ref{sec:algorithm} demonstrates the construction of the calendar layout in detail. Section \ref{sec:opt} lists and describes the options that come with the \code{frame_calendar} function. Section \ref{sec:examples} presents some variations of its usage. Section \ref{sec:discussion} discusses the advantages and disadvantages of the method.

@@

Figure \ref{fig:flinders-2016} shows the line glyphs organised in the monthly calendar over the year of 2016. This is achieved by the \code{frame_calendar} function computing the new coordinates according to the input data variables; in turn the rearranged data values are plotted using the \pkg{ggplot2} package [@R-ggplot2], which is the implementation of the grammar of graphics (@wilkinson2006grammar and @wickham2010layered).

@@

# Options
\label{sec:opt}

The \code{frame_calendar} function assumes that the underlying data structure is tidy data [@wickham2014tidy].

```
frame_calendar(
  data, x, y, date, calendar = "monthly", dir = "h", sunday = FALSE, 
  nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed"
)
```

@@

The comparison of one sensor to another adds additional insights to the dataset, which is commonly done with an overlaying plot such as Figure \ref{fig:overlay}. For instance, the dominant patterns occurred to both train stations--Flinders Street Station and Flagstaff station--are together driven by the commuters on the work days; however, the former is overwhelmed by the number of pedestrians during the weekends and public holidays, compared to the latter. It suggests that Flagstaff station limits its functionality as simply the public transport hub for day-to-day commuters, but various activities take place around the Flinders Street Station other than commuting. Because Flinders Street Station closely lies to South Bank and other places of attractions where the locals go for leisure and the tourists for travelling. It can be noted from Figure \ref{fig:overlay} that State Library follows the similar temporal trend as Flinders Street Station on non-work days.

```{r overlay, fig.height = 6.5, fig.cap = "Superimposing line graphs of three sensors in the monthly calendar. The orange line and the purple line, represented by the number of pedestrians at Flagstaff Station and Flinders Street Station, are on top of the green one indicating State Library. Flagstaff station is not as busy as the other two on non-work days."}
```

The nighttime events, such as White Night and New Year Evening Firework, has barely affected the operation of Flagstaff Station but heavily affected the incoming traffic to Flinders Street Station and State Library on these days.

@@

```{r scatterplot, fig.height = 5, fig.cap = "Lag scatterplot in the daily calendar. Lagged hourly counts are plotted against current ones to demonstrate the autocorrelation at the order of 1. The positive association between the one-hour-ahead count and the present regardless of the data values takes place on non-working days."}
```

### 2017-07-11 main.tex
A few tweaks

\Abstract{
It is challenging to visualise people's daily schedules as such data are
often of multiple seasons and of long time span. One example is
Melbourne pedestrian sensor data which are collected at hourly interval
and at a number of locations. Owing to human activities, the data
features multiple temporal scales such as time of day, day of week, day
of year. We develop the \code{frame\_calendar} function that provides a
method for rearranging such kind of data in a calendar-based format and
accomplishing visualisation with the grammar of graphics, and describe
its construction in detail. We provide some examples to demonstrate its
usage by applying to the pedetrian data.
}

@@

This work was originally motivated by studying foot traffic in the city
of Melbourne \citep{ped}. There have been 43 installed sensors counting
pedestrians every hour across the downtown area until the end of 2016
(see Figure \ref{fig:ped-map}). The dataset can shed lights into
understanding people's daily schedules, or assisting administration and
business planning. We start off with the conventional time series plot
to take a glimpse at such data. Small multiples, shown as Figure
\ref{fig:time-series-plot}, give an overall picture of foot traffic at a
selection of 7 sensors. Figure \ref{fig:facet-time}, complemented to the
Figure \ref{fig:time-series-plot}, provides another aspect of temporal
patterns in the more detailed level by faceting day of the week. It
lends itself to a number of issues that make exploratory data
visualisation challenging in many temporal-context applications
involving human behaviours:

@@

\caption[Time series plot about the number of pedestrians over the year of 2016 at a selection of 7 sensors in the city of Melbourne]{Time series plot about the number of pedestrians over the year of 2016 at a selection of 7 sensors in the city of Melbourne. Coloured by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. The traffic, specifically at two train stations (Flagstaff Station and Flinders Street Station), is primarily driven by the commuters, as depicted in the graph.}\label{fig:time-series-plot}
\end{figure}
\end{CodeChunk}

@@

Calendar-based graphics turn out to be a useful tool in unfolding
human-related activities in temporal context (cite). For example,
\citet{VanWijkCluster1999} developed a calendar view of heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It remarkably
contrasts weekdays and weekends, highlights public holiday, and presents
other known seasons like school vacations, all of which have influence
over the turn-outs in the office. The calendar-based heatmap was
implemented in a couple of R packages: \pkg{ggTimeSeries}
\citep{R-ggTimeSeries} and \pkg{ggcal} \citep{R-ggcal}. However, these
techniques are too constrained to colour-encoding graphics and day of
week as the smallest time scale. Time of day, which serves as one of the
most important aspect in explaining variations arising from pedestrian
sensor data, will be neglected through daily aggregation. Additionally
if simply using colours encoded as quantities, it may becomes less
perceivable of the actual variations or behaviours from the viewer's
perspective.

@@

Figure \ref{fig:flinders-2016} shows the line glyphs organised in the
monthly calendar over the year of 2016. This is achieved by the
\code{frame_calendar} function computing the new coordinates according
to the input data variables; in turn the rearranged data values are
plotted using the \pkg{ggplot2} package \citep{R-ggplot2}, which is the
implementation of the grammar of graphics (\citet{wilkinson2006grammar}
and \citet{wickham2010layered}). The \code{frame_calendar} function
assumes that the underlying data structure is tidy data
\citep{wickham2014tidy}.

@@

@@

The comparison of one sensor to the other adds additional insights to
the dataset, which is commonly done with an overlaying plot such as
Figure \ref{fig:overlay}. For instance, the dominant patterns occurred
to both train stations--Flinders Street Station and Flagstaff
station--are together driven by the commuters on the work days; however,
the former is overwhelmed by the number of pedestrians during the
weekends and public holidays, compared to the latter. It suggests that
Flagstaff station limits its functionality as simply the public
transport hub for day-to-day commuters, but various activities take
place around the Flinders Street Station other than commuting. Because
Flinders Street Station closely lies to South Bank and other places of
attractions where the locals go for leisure and the tourists for
travelling.

@@

\caption[Superimposing line graphs of two sensors on the monthly calendar]{Superimposing line graphs of two sensors on the monthly calendar. The orange line, represented by the number of pedestrians at Flinders Street Station, is on top of the green one indicating Flagstaff Station. Flinders Street Station attracts more pedestrians during the afternoon time and non-working days.}\label{fig:overlay}
\end{figure}
\end{CodeChunk}

The nighttime events, such as White Night and New Year Evening Firework,
has barely affected the operation of Flagstaff Station but heavily
affected the incoming traffic to Flinders Street Station on these days.

>>>>

\Abstract{
It is challenging to visualise people's daily schedules as such data are
often of multiple seasons and of long time span. One example is
Melbourne pedestrian sensor data which are collected at hourly interval
and at a number of locations. Owing to human activities, the data
features multiple temporal scales such as time of day, day of week, day
of year. We develop the \code{frame\_calendar} function that provides a
method for rearranging such kind of data in a calendar-based format and
accomplishing visualisation with the grammar of graphics. We describe
its construction along with a few options. We provide some examples to
demonstrate its usage by applying to the pedetrian data.
}

@@

This work was originally motivated by studying foot traffic in the city
of Melbourne \citep{ped}. There have been 43 installed sensors counting
pedestrians every hour across the downtown area until the end of 2016
(see Figure \ref{fig:ped-map}). The dataset can shed lights into
understanding people's daily schedules, or assisting administration and
business planning. We start off with the conventional time series plot
to take a glimpse at such data. Small multiples, shown as Figure
\ref{fig:time-series-plot}, give an overall picture of foot traffic at a
selection of 3 sensors. Figure \ref{fig:facet-time}, complemented to the
Figure \ref{fig:time-series-plot}, provides another aspect of temporal
patterns in the more detailed level by faceting day of the week. It
lends itself to a number of issues that make exploratory data
visualisation challenging in many temporal-context applications
involving human behaviours:

@@

\caption[Time series plot about the number of pedestrians over the year of 2016 at a selection of 3 sensors in the city of Melbourne]{Time series plot about the number of pedestrians over the year of 2016 at a selection of 3 sensors in the city of Melbourne. Coloured by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. The traffic, specifically at two train stations (Flagstaff Station and Flinders Street Station), is primarily driven by the commuters, as depicted in the graph.}\label{fig:time-series-plot}
\end{figure}
\end{CodeChunk}

@@

Calendar-based graphics turn out to be a useful tool in unfolding
human-related activities in temporal context. For example,
\citet{VanWijkCluster1999} developed a calendar view of heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It remarkably
contrasts weekdays and weekends, highlights public holiday, and presents
other known seasons like school vacations, all of which have influence
over the turn-outs in the office. The calendar-based heatmap was
implemented in a couple of R packages: \pkg{ggTimeSeries}
\citep{R-ggTimeSeries} and \pkg{ggcal} \citep{R-ggcal}. However, these
techniques are too constrained to colour-encoding graphics and day of
week as the smallest time scale. Time of day, which serves as one of the
most important aspect in explaining variations arising from pedestrian
sensor data, will be neglected through daily aggregation. Additionally
if simply using colours encoded as quantities, it may becomes less
perceivable of the actual variations or behaviours from the viewer's
perspective.

@@

Figure \ref{fig:flinders-2016} shows the line glyphs organised in the
monthly calendar over the year of 2016. This is achieved by the
\code{frame_calendar} function computing the new coordinates according
to the input data variables; in turn the rearranged data values are
plotted using the \pkg{ggplot2} package \citep{R-ggplot2}, which is the
implementation of the grammar of graphics (\citet{wilkinson2006grammar}
and \citet{wickham2010layered}).

@@

\label{sec:opt}

The \code{frame_calendar} function assumes that the underlying data
structure is tidy data \citep{wickham2014tidy}.

\begin{verbatim}
frame_calendar(
  data, x, y, date, calendar = "monthly", dir = "h", sunday = FALSE, 
  nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed"
)
\end{verbatim}

@@

The comparison of one sensor to another adds additional insights to the
dataset, which is commonly done with an overlaying plot such as Figure
\ref{fig:overlay}. For instance, the dominant patterns occurred to both
train stations--Flinders Street Station and Flagstaff station--are
together driven by the commuters on the work days; however, the former
is overwhelmed by the number of pedestrians during the weekends and
public holidays, compared to the latter. It suggests that Flagstaff
station limits its functionality as simply the public transport hub for
day-to-day commuters, but various activities take place around the
Flinders Street Station other than commuting. Because Flinders Street
Station closely lies to South Bank and other places of attractions where
the locals go for leisure and the tourists for travelling. It can be
noted from Figure \ref{fig:overlay} that State Library follows the
similar temporal trend as Flinders Street Station on non-work days.

@@

\caption[Superimposing line graphs of three sensors in the monthly calendar]{Superimposing line graphs of three sensors in the monthly calendar. The orange line and the purple line, represented by the number of pedestrians at Flagstaff Station and Flinders Street Station, are on top of the green one indicating State Library. Flagstaff station is not as busy as the other two on non-work days.}\label{fig:overlay}
\end{figure}
\end{CodeChunk}

The nighttime events, such as White Night and New Year Evening Firework,
has barely affected the operation of Flagstaff Station but heavily
affected the incoming traffic to Flinders Street Station and State
Library on these days.
